Latin epigraphy for the visually impaired. New technologies to favour universal accessibility

### 2015-12-16 source/9/Latin epigraphy for the visually impaired.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Francesca Licordari}
\date{2015-11-19}
\begin{document}
1. Latin epigraphy for the visually impaired.

\ \ \ \ New technologies to favour universal accessibility

\bigskip

Francesca Licordari

Museo Nazionale Etrusco di Villa Giulia, Roma

francesca.licordari@beniculturali.it

\bigskip

Abstract

\bigskip

The problem of accessibility to works of art for the visually impaired can be extended to a system such as Latin
epigraphy, which presents great difficulties since the engraved letters are not clear to the touch. The technology of
3D printing allows the creation of materials that make it easier to read epigraphic texts in following the “design for
all” principle demonstrating an economic, rapid technique that is easy to implement and reasonably robust.

\bigskip

Keywords:

\bigskip

Accessibility, visually impaired, 3D printing, design for all, Braille, epigraphy

\bigskip

1.1. Introduction

\bigskip

The Declaration of the Rights of Deaf-Blind Persons, adopted by the UN in 1979, provides that every blind or deaf person
has the right to enjoy the same privileges, guaranteed to all people and to have their aspirations and abilities
recognized and respected. These principles are reaffirmed by the UN Convention on the Rights of Persons with
Disabilities of 2006, which states (Art. 30) that the Member States must recognize “the right of persons with
disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
ensure to persons with disabilities:

(A) enjoy access to cultural materials in accessible formats;

(B) enjoy access to television programs, films, theater and other cultural activities, in accessible formats;

(C) enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
and, as far as possible, to monuments and sites of national cultural importance.”

@@

The bibliography in this area of study is unfortunately not very comprehensive, especially as regards the epigraphic
field; but advances are even now being made. In Italy, the most important studies are conducted by the “Museo Tattile
Statale Omero” in Ancona, which, being a museum of statuary, is mainly engaged with reproductions and the accessibility
of tactile sculptures. The same work, involving the reproduction of paintings, is being carried out by the “Museo
Tattile Anteros” in Bologna. It is no coincidence that Italian studies in this field are the most important done at the
international level. This is mainly due to the fact of Italy’s great artistic and epigraphic heritage, which makes it
the country where the problem is most profoundly felt.

@@

\bigskip

\bigskip

1.2. Problems in Understanding Epigraphy

\bigskip

The field of epigraphy, always considered highly specialized, can be of great fascination to people with visual
impairment. Some may have studied ancient languages at school, especially Latin, and so feel the need to have practical
examples for the understanding of the subject; and some may be attracted out of curiosity. It is obvious that
accessibility to the inscriptions for such people involves a number of difficulties, and not just that of understanding
the texts, which are often full of abbreviations. There is, as well, a certain percentage of the blind who never
learned to read and write. Not infrequently, they may only know capitals, but not cursive characters. The inscriptions,
which appear so sharp to us that we are often tempted to touch them with our hands to follow the course of the
incisions, actually create complicated problems for the blind.

@@

The best solution to make the epigraphic media available to the visually impaired turns out to be, therefore, one that
involves more visitors who can better understand the artwork. In this regard, it being understood that the chance of
touching an original work wherever possible is a uniquely fascinating act, it becomes necessary to propose
reproductions that can make the material accessible by following the concept of “design for all,” which is adaptable to
all types of art. As rightly noted by the scholar and educator Enzo Tioli (Tioli, 2006 in L’arte a portata di mano),
this principle can be defined as “identical, whenever possible, equivalent when that is not possible.” All that can be
touched and seen in the original work constitutes the “identical”, the use of a copy constitutes the “equivalent”.

\bigskip

\bigskip

1.3. The Techniques of Accessibility

\bigskip

In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
and also to replace those exhibits that had been located outside and brought indoors to protect them from weather
conditions. These casts, if used by the visually impaired, present the same problems of accessibility as the original
in terms of the difficulty of understanding the incision’s grooves. It is necessary, therefore, to have a reproduction
of a different type, upon which the letters can be transliterated in relief.

@@

A technique which solves this problem is that of the Thermoform, that is a relief obtained from a plastic material,
formed in a metal matrix. This type of technique is the one most used for the creation of teaching aids in schools for
the blind (in this regard see www.prociechi.it). But the technique proves to be very expensive in the creation of the
original matrix, as well as for the reproductions of individual pieces. The costs come down significantly when multiple
copies are derived. The ease of obtaining new copies compensates for the rapid deterioration of the material.

@@

\bigskip

\bigskip

1.4. 3D Printing

\bigskip

The new frontier in the understanding of works of art for the sight impaired is provided by 3D printing, which proves to
be both economical, fast, easy to implement, and durable. It is a single technique, which gathers together the
advantages of the techniques already highlighted and overcomes the defects of poor durability. Moreover, as the name
implies, it allows the introduction of the third dimension, which until now had never been possible, allowing the blind
person to be able to better perceive objects in space. This problem maybe not obvious in the case of inscriptions on
plates, but presents itself when we deal with sculptured monuments containing inscriptions.

Precisely from this technology in 2007, coordinated by Prof. Massimo Bergamasco and the Perceptual Robotics Laboratory
(Percro), was developed the project PURE-FORM, which led to the creation of a virtual museum, which gives visitors the
chance to see and “touch” objects, in particular digital sculptures, belonging to different historical periods and from
different contexts. All of this is based on a digital scan of the artworks stored in a database, thus offering a wide
choice, and reproduced on a touch screen display. In this project the focus was also placed exclusively on sculptures,
which already possess the third dimension, while the epigraphic content has been completely omitted. The creation of a
display, in practice a computer, allows on the one hand more accessibility to artworks, but on the other deprives the
user of the opportunity to travel to a place where they could personally experience the work’s features. It must not be
forgotten that a museum or an archaeological site are not just warehouses and storage sites.

@@

{}- Original

{}- Original with fractures and abrasions and signs of aging

{}- Integrated original

{}- Reproduction with transliterated reliefs

{}- Reproduction of the original with Braille overlay.

@@

A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy (Licordari, 2015), which
supports an exhibition project for the dissemination of Latin Epigraphy to the public, proposes 3D reproduction of two
mosaics of the stationes of that square in Ostia, with images and writing – that of the Libyan boatmen of Sabratha and
of the Carthaginian boatmen. The first, with the inscription STAT. SABRATENSIVM [CIL 14, 4549.14], is characterized by
the image of an elephant, the symbol of the Sabratha ivory trade. The second contains the inscription -NAVICVL.
KARTHAG. DE SVO [CIL 14, 4549.18] along with two ships. The paper’s original idea was based on the fact that not only
should there be brief and easy to understand inscriptions, but that the understanding could be made easier by the
presence of drawings presumably known to blind people. Certainly the case of Ostia Antica is a particularly simple one
as it involves mosaics of only two colors. A more complicated case would be a polychrome mosaic which attempts to
assimilate all the aspects of a colored painting. It is obvious that the sketched designs and writing can be configured
and presented on different levels.

@@

\bigskip

\bigskip

1.5. Conclusions

\bigskip

As can be seen, the studies in this field as well as the examples of works on which to experiment with seeing impaired
users are as yet few in number. We hope that this essay will lead to inspire a dialogue among the relevant experts, and
that more work will be done in the field of epigraphy to make it increasingly accessible for persons with visual
disabilities.

@@

\bigskip

Bibliography:

\bigskip

AA.VV., L’accessibilità al patrimonio museale per le persone con minorazione visiva, CD-ROM Le dispense del Museo Omero,
Ancona 2014

AA.VV., L’arte a portata di mano. Verso una pedagogia di accesso ai Beni Culturali senza barriere, Atti del convegno
Portonovo di Ancona 21-23 ottobre 2004, Roma 2006

Bellini, A., Toccare l’Arte. L’educazione estetica di ipovedenti e non vedenti, Roma 2000

Cetorelli Schivo, G., Colcerasa, M., Disabilità e museo. Percorrenze culturali di emergenza, in “Museo e territorio”,
Atti della III giornata di studi (Velletri 7-8 marzo 2003), Roma 2004

Hersh, M, Johnson, M.A. (Eds.), Assistive Technology for Visually Impaired and Blind People, Glasgow 2008

Jansson, G., Bergamasco, M., Frisoli, A., A new option for the visually impaired to experience 3D art at museums: manual
exploration of virtual copies, in “Visual Impairment Research”, 5,1, 2003, pp. 1–12

Licordari, F., Tituli in artis operibus inscripti: proposta di allestimento di una mostra nei locali della Fondazione
Sorgente, tesi di laurea della Scuola di Specializzazione in Beni Archeologici, Roma 2015

Levi, F., L’accessibilità alla cultura per i disabili visivi. Storia e orientamenti, Torino 2015

Velestino, D., Simili, G., Poscolieri, M. (Eds.), Messaggi dalla pietra, Torino 2007
\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Latin epigraphy for the visually impaired. New technologies to favour universal accessibility}
\author[mne]{Francesca Licordari\corref{first}}
\address[mne]{Museo Nazionale Etrusco di Villa Giulia, Roma}
\cortext[first]{Corresponding author. Email: francesca.licordari@beniculturali.it}
\date{2015-11-19}
\begin{document}
\maketitle
\begin{abstract}
The problem of accessibility to works of art for the visually impaired can be extended to a system such as Latin
epigraphy, which presents great difficulties since the engraved letters are not clear to the touch. The technology of
3D printing allows the creation of materials that make it easier to read epigraphic texts in following the ``design for
all'' principle demonstrating an economic, rapid technique that is easy to implement and reasonably robust.
\end{abstract}
\keywords{Accessibility, visually impaired, 3D printing, design for all, Braille, epigraphy}

\section{Introduction}
\noindent The Declaration of the Rights of Deaf-Blind Persons, adopted by the UN in 1979, provides that every blind or deaf person
has the right to enjoy the same privileges, guaranteed to all people and to have their aspirations and abilities
recognized and respected. These principles are reaffirmed by the UN Convention on the Rights of Persons with
Disabilities of 2006, which states (Art. 30) that the Member States must recognize ``the right of persons with
disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
ensure to persons with disabilities:

\begin{description}
\item[A] enjoy access to cultural materials in accessible formats;

\item[B] enjoy access to television programs, films, theater and other cultural activities, in accessible formats;

\item[C] enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
and, as far as possible, to monuments and sites of national cultural importance.''

\end{description}

@@

The bibliography in this area of study is unfortunately not very comprehensive, especially as regards the epigraphic
field; but advances are even now being made. In Italy, the most important studies are conducted by the ``Museo Tattile
Statale Omero'' in Ancona, which, being a museum of statuary, is mainly engaged with reproductions and the accessibility
of tactile sculptures. The same work, involving the reproduction of paintings, is being carried out by the ``Museo
Tattile Anteros'' in Bologna. It is no coincidence that Italian studies in this field are the most important done at the
international level. This is mainly due to the fact of Italy’s great artistic and epigraphic heritage, which makes it
the country where the problem is most profoundly felt.

@@

\section{Problems in Understanding Epigraphy}
\noindent The field of epigraphy, always considered highly specialized, can be of great fascination to people with visual
impairment. Some may have studied ancient languages at school, especially Latin, and so feel the need to have practical
examples for the understanding of the subject; and some may be attracted out of curiosity. It is obvious that
accessibility to the inscriptions for such people involves a number of difficulties, and not just that of understanding
the texts, which are often full of abbreviations. There is, as well, a certain percentage of the blind who never
learned to read and write. Not infrequently, they may only know capitals, but not cursive characters. The inscriptions,
which appear so sharp to us that we are often tempted to touch them with our hands to follow the course of the
incisions, actually create complicated problems for the blind.

@@

The best solution to make the epigraphic media available to the visually impaired turns out to be, therefore, one that
involves more visitors who can better understand the artwork. In this regard, it being understood that the chance of
touching an original work wherever possible is a uniquely fascinating act, it becomes necessary to propose
reproductions that can make the material accessible by following the concept of ``design for all,'' which is adaptable to
all types of art. As rightly noted by the scholar and educator Enzo Tioli \citep{Tioli2006},
this principle can be defined as ``identical, whenever possible, equivalent when that is not possible.'' All that can be
touched and seen in the original work constitutes the ``identical'', the use of a copy constitutes the ``equivalent''.

\section{The Techniques of Accessibility}
\noindent In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
and also to replace those exhibits that had been located outside and brought indoors to protect them from weather
conditions. These casts, if used by the visually impaired, present the same problems of accessibility as the original
in terms of the difficulty of understanding the incision’s grooves. It is necessary, therefore, to have a reproduction
of a different type, upon which the letters can be transliterated in relief.

@@

A technique which solves this problem is that of the Thermoform, that is a relief obtained from a plastic material,
formed in a metal matrix. This type of technique is the one most used for the creation of teaching aids in schools for
the blind (in this regard see \url{www.prociechi.it}). But the technique proves to be very expensive in the creation of the
original matrix, as well as for the reproductions of individual pieces. The costs come down significantly when multiple
copies are derived. The ease of obtaining new copies compensates for the rapid deterioration of the material.

@@

\section{3D Printing}
\noindent The new frontier in the understanding of works of art for the sight impaired is provided by 3D printing, which proves to
be both economical, fast, easy to implement, and durable. It is a single technique, which gathers together the
advantages of the techniques already highlighted and overcomes the defects of poor durability. Moreover, as the name
implies, it allows the introduction of the third dimension, which until now had never been possible, allowing the blind
person to be able to better perceive objects in space. This problem maybe not obvious in the case of inscriptions on
plates, but presents itself when we deal with sculptured monuments containing inscriptions.

Precisely from this technology in 2007, coordinated by Prof. Massimo Bergamasco and the Perceptual Robotics Laboratory
(Percro), was developed the project PURE-FORM, which led to the creation of a virtual museum, which gives visitors the
chance to see and ``touch'' objects, in particular digital sculptures, belonging to different historical periods and from
different contexts. All of this is based on a digital scan of the artworks stored in a database, thus offering a wide
choice, and reproduced on a touch screen display. In this project the focus was also placed exclusively on sculptures,
which already possess the third dimension, while the epigraphic content has been completely omitted. The creation of a
display, in practice a computer, allows on the one hand more accessibility to artworks, but on the other deprives the
user of the opportunity to travel to a place where they could personally experience the work’s features. It must not be
forgotten that a museum or an archaeological site are not just warehouses and storage sites.

@@

\begin{itemize}
\item Original

\item Original with fractures and abrasions and signs of aging

\item Integrated original

\item Reproduction with transliterated reliefs

\item Reproduction of the original with Braille overlay.
\end{itemize}

@@

A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy \citep{Licordari2015}, which
supports an exhibition project for the dissemination of Latin Epigraphy to the public, proposes 3D reproduction of two
mosaics of the stationes of that square in Ostia, with images and writing – that of the Libyan boatmen of Sabratha and
of the Carthaginian boatmen. The first, with the inscription STAT. SABRATENSIVM [CIL 14, 4549.14], is characterized by
the image of an elephant, the symbol of the Sabratha ivory trade. The second contains the inscription -NAVICVL.
KARTHAG. DE SVO [CIL 14, 4549.18] along with two ships. The paper’s original idea was based on the fact that not only
should there be brief and easy to understand inscriptions, but that the understanding could be made easier by the
presence of drawings presumably known to blind people. Certainly the case of Ostia Antica is a particularly simple one
as it involves mosaics of only two colors. A more complicated case would be a polychrome mosaic which attempts to
assimilate all the aspects of a colored painting. It is obvious that the sketched designs and writing can be configured
and presented on different levels.

@@

\section{Conclusions}
\noindent As can be seen, the studies in this field as well as the examples of works on which to experiment with seeing impaired
users are as yet few in number. We hope that this essay will lead to inspire a dialogue among the relevant experts, and
that more work will be done in the field of epigraphy to make it increasingly accessible for persons with visual
disabilities.

@@

\nocite{accessibilità,Bellini2000,Cetorelli2004,Hersh2008,Jansson2003,Levi2015,Velestino2007}

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}
